---
title: "R workshop #1: plotting and manipulating data"
author: "Nicola Romanò"
#date: 05 October 2018
output: 
  tufte::tufte_handout: default
  tufte::tufte_html: default
---

```{r setup, include=FALSE}
library(tufte)
library(xtable)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(global.par = TRUE)

# See https://stackoverflow.com/questions/25646333/code-chunk-font-size-in-rmarkdown-with-knitr-and-latex
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) 
  {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n\\normalsize"), x)
  })
```


***
# Introduction

Plotting is an extremely important part of data analysis. Indeed, after you finish collecting a dataset one of the first thing you should probably do is plotting the data. This allows you to get a feel for what your data look like and to spot obvious errors in their collection, or with the experimental setup. However, there is so much more that a good graph can do: it can help you support your choice of statistical test used for analysing those data, it can help you better understand the underlying causes of the phenomenon that you are observing, and help you explain those to others.
This process is so important that John W. Tukey, one of the fathers of modern statistics coined the phrase _Observational Data Analysis_ (ODA) to describe it ^[Tukey, Observational Data Analysis, 1977].

Data quality is of course very important. Remember the phrase "garbage in, garbage out"! If you collect data in a "sloppy" manner you cannot expect to get beautiful and reproducible results. Similarly, if you do not include the appropriate controls you may have some wonderful data... that means nothing! The way you structure your data is also an extremely important issue to consider. Certain formats may be easier to read by a human, but won't be easy for a computer to process (remember what learnt in the lectures). It has been estimated that between 30% and 80% of data analysis consists of data preparation ^[Dasu and Johnson 2003]. This is something that you will encounter over and over again in your career as a biomedical researcher, so it is extremely important to get into _good habits_ of data collection as early as possible.

# Learning objectives
After completing this workshop you will be able to:

* Reading and explore the main features of a dataset
* Identify suspicious data points / outliers
* Plot the data and visually identify relationships between variables
* Convert wide dataset into long datasets

# Basic data handling & plotting using R
This first section will be a refresher on how to read, explore and plot data in R. You should be fairly familiar with these commands, but a refresher always helps!

For this first part we are going to use the file called _metab-workshop1.csv_.

Let's start by loading the dataset using^[You will have to change the path to where you put the file, e.g. _read.csv("C:/Workshop1/metab-workshop1.csv")_. Alternatively you can tell R to look for files in that directory using the _setwd_ command. This second approach is very useful if you are reading/writing multiple files in the same script.]

```{r}
metab <- read.csv("metab-workshop1.csv")
```
This dataset contains the concentration of a metabolite in the plasma of patients after receiving a specific treatment. We also have information about the patient sex and age. The first thing we should do is examine our data.

_head_ allows us to look at the first few lines of the dataset. Simply use it as follows ^[Try to use the _n_ parameter to see a specific number of lines. If you want to see the end of your dataset you can use _tail_ instead of _head_.]:

```{r}
head(metab)
```

The dataset shows the data for each patient in a row, and has `r ncol(metab)` columns showing the measured values or characteristics for that specific patient. We can use _colnames_ to see the name of the columns in our dataset.^[Unsurprisingly, there is a _rownames_ function as well. Try it!]

```{r}
colnames(metab)
```

We can now check how many patients we have. Since each line of our dataset represents one patient we can simply count the number of rows using _nrow_ ^[Can you guess how to count columns instead?].

```{r}
nrow(metab)
```

Suppose we wanted to know how many of those `r nrow(metab)` patients are men and how many are women. There are a few ways of doing this in R. 
For example you can subset your data and then count the rows.^[Remember, R uses square brackets to access the content of a dataset (the correct R name for what you are using is a _data frame_). For example writing _data[1,5]_ will get the element in row 1, column 5. Using _data[1,]_ will get row 1 and all of the columns.]

```{r}
# Take all rows for which the _Sex_ = "F" and all of the columns.
females <- metab[metab$Sex == "F",] 
nrow(females)
nrow(metab) - nrow(females) # Number of men = total - women
```

A more elegant way to do this is to use the _table_ function.

```{r}
table(metab$Sex)
```

You can pass multiple parameters to table (separated by ,). Try creating a table of Sex and Treatment^[Note that the order in which you pass parameters to _table_ is important! Try passing _metab\$Sex_ either before or after _metab\$Treatment_. What happens?]. 

How many men are there in the control group? _____________

Another neat way to explore the data is to use the _summary_ function:

```{r size="small"}
summary(metab)
```

This tells us a lot! We have four variables (Concentration, Sex, Age and Treatment). Concentration is a continuous variable, so we get statistics about its distribution. Sex is a discrete variable (in R these are called factors) and has two possible levels: F and M.^[R orders levels alphabetically, we will see how to change this later on.]
What type of variables are Age and Treatment? _______________________________

\vspace{2em}

Do you notice any issue with these data? ________________________
________________________________________________________________

\vspace{1em}

Can you think of a way of correcting the problem? ________________
________________________________________________________________


Obviously you can also get summary statistics using R functions such as _mean_, _median_, _range_, _min_, _max_, _quantile_^[By default the quantile function returns 5 quantiles (min, 25%, median, 75%, max), but try specifying the _probs_ parameter, for example as _quantile(metab$Age, probs = c(0.3, 0.6))_. What happens?]

For example
```{r}
median(metab$Age)
```

# Basic plotting

We now want to start plotting the values we read.
```{r echo = FALSE}
par(mar = c(5, 4, 1, 2))
```

```{r fig.height=2.5}
plot(metab$Concentration)
```

This plot shows the `r nrow(metab)` concentration measurements in our dataset, and easily allows us to estimate their mean^[Could you guess the mean of the values by looking at the plot? Use the _mean_ function to calculate their actual mean. Were you close?].
Aside from this, it is not particularly useful, is quite ugly and cluttered. Let's make it prettier!
\newpage
```{r fig.height=2.5}
plot(metab$Concentration, 
     pch = 20, cex = 0.7, cex.axis = 0.8, las = 1, 
     ylab = "Concentration")
```
We have added several parameters: 

- _pch_ changes the *p*lotting *ch*aracter^[There are many different plotting characters, try different numbers for _pch_ and see what happens. You can even use a character of your choice, e.g. _pch='x'_.]
- _cex_ changes the size of the plotting points. Values less then 1 make them smaller, values bigger than 1 make them bigger. Depending on the configuration of your computer, different values may give prettier results.
- _cex.axis_ changes the size of the axes labels (both x and y).
- _ylab_ sets the label for the y axis^[Try changing the label of the x axis].
- Try changing the _las_ parameter and see what it does.

There are many more parameters that you can use, we are going to see a few, but if you want to see the full list type _help(par)_^[_par_ is an extremely powerful function, it allows you to set the default plotting parameters for all of the graphs you will plot after calling it. It also allows you to have multiple plots in the same figure, using the _mfrow_ parameter. Try for example writing _par(mfrow = c(1, 3))_, then doing 3 different plots.
Can you plot concentration in men on the top and women on the bottom? Pay attention to the y axis scale!].

\vspace{1em}

Now that the plot looks prettier, let's try to make it more useful! We can color the points depending on the value in the Sex column. Let's create a new variable, that contains `r nrow(metab)` elements, one for each sample, equal to _"darkgreen"_ if the sample is from a man and _"purple"_ if it is from a woman.^[Curious to know which colours are available in R? Type _colours()_!]. We assign _"darkgreen"_ to everyone, using the _rep_ function^[The _rep_ function repeats a value a specific number of times. For example, try, _rep(5, 8)_. Note the quotes around the values, indicating that these are not names of variables but the actual values we want.], then change to _"purple"_ the values corresponding to samples from women. 

\newpage
```{r fig.height=3}
pointcolor <- rep("darkgreen", nrow(metab))
pointcolor[metab$Sex == "F"] <- "purple"
plot(metab$Concentration, col = pointcolor, bty = "n",
     pch = 20, cex = 0.7, cex.axis = 0.8, las = 1, 
     ylab = "Concentration")
legend("topright", legend = c("Men", "Women"), fill = c("darkgreen", "purple"), cex = 0.7)
```

By this time, you should have seen that there is a negative value of concentration! Since this is clearly an error^[Why is this?], we can safely remove it.

```{r}
# This returns the number of the row(s) containing
# negative values, in this case row 49
wrong.val <- which(metab$Concentration < 0)
# Using - in front of the row (or column) number removes that row (or column). By re-assigning the result to metab we are overwriting its previous value
metab <- metab[-wrong.val,]
```

# Grouping data

Plotting single data points is useful, but most often we want to group the plot by some meaningful factor. Boxplots (a.k.a. box-and-whisker plots) are very good for this.

\newpage
```{r echo = FALSE}
par(cex.axis = 0.8)
```

```{r fig.height=2.5}
boxplot(Concentration ~ Treatment, metab, las = 1, 
        ylim = c(0, 1000), pch = 20,
        ylab = "Concentration")
```

The _boxplot_ function allows us to plot a value against one or more factor^[Why is there a point on top of the boxplot for group A? Try and find out, we can discuss this in the forum!]. As we noted above, R orders factors alphabetically, therefore groups appear in the _odd_ order A, B, and Control. Should we want to change the order we could do:

```{r}
levels(metab$Treatment) <- c("CTRL", "A", "B")
```

After doing this, the boxplot will be correctly ordered (try it!).

We can also combine more than one factor in the boxplot, using + (e.g. Concentration ~ Treatment + Sex)

Try plotting concentration against treatment, separately for men and women. 

Now try swapping Sex and Treatment, what happens?

Sometimes it is useful to plot single points over the boxplot. This can be done using the _stripchart_ function.

```{r}
boxplot(Concentration ~ Treatment, metab, pch = 20)
stripchart(Concentration ~ Treatment, metab, add = TRUE, vert = TRUE, 
           pch = 20, cex = 0.6, 
           method = "jitter", jitter = 0.1)
```

We need to pass two important parameters to _stripchart_: _add_ tells R to plot over the existing graph, rather than producing a new one. _vert_ specifies to plot the stripchart vertically and not horizontally ^[The _jitter_ parameter moves points around and is very handy when you have a lot of overlapping points. It only works when specifying _method="jitter"_. Try changing the _jitter_ value and see what happens!].

# Histograms and density plots
Histograms are a great way to summarise data. Let's generate an histogram of the concentration in our dataset.

```{r}
hist(metab$Concentration, col = "black", las = 1,
     main = "Concentration of metabolite", xlab = "Concentration")
```

The _br_ parameter can be used to specify the number of columns (the 'breaks') in the histogram^[By default R uses Sturge's formula to determine the number of columns. There are many different ways of calculating this number. Wikipedia has a fairly lengthy section about number of bins here: https://en.wikipedia.org/wiki/Histogram]. It can be used in two ways: by specifying the number of breaks or by specifying the values at which we want to break the histogram.
\newpage
For instance:

```{r}
# Histogram with 100 columns
hist(metab$Concentration, col = "black", las = 1, br = 100,
     main = "Concentration of metabolite", xlab = "Concentration")
# Histogram with columns from 0 to 1000 with width 50 
hist(metab$Concentration, col = "black", las = 1, br = seq(0, 1000, 50),
     main = "Concentration of metabolite", xlab = "Concentration")
```

Which is the best? Let's discuss this in the forum!

# Other plots

There are many other plots that R is able to produce, including some very complex ones. Some plots require extra packages (such as _ggplot2_) to be installed. We will not cover those here, but there are many online resources that you can use to learn about them should you want to.

What does this code do? Is the resulting plot more or less useful than a boxplot? Why? Try to run it, change things, read the R help, and then we can discuss your solution together!

```{r eval=FALSE}
males <- metab$Concentration[metab$Sex == 'M']
females <- metab$Concentration[metab$Sex == 'F']

mean.M <- mean(males) 
mean.F <- mean(females) 
sd.M <- sd(males)
sd.F <- sd(females)
# barplots returns the x position of each bar
bp <- barplot(c("Men" = mean.M, "Women" = mean.F), col = c("green", "orange"), ylim = c(0, 600), las = 1, ylab = "Concentration")
arrows(bp, c(mean.M - sd.M, mean.F - sd.F), bp, c(mean.M + sd.M, mean.F + sd.F),
       angle = 90, code = 3)  
```

# Converting between wide and long data
The last part of this workshop concerns data formatting. You have learnt about long and wide data formatting during the lectures. Very often you will have to deal with _messy_ data, that is not formatted in a way that is easy to work with (e.g. data may be in wide format). This may be because someone else recorded the data, or because often wide data is easier to look at in software such as Excel.
Read the file _lizard.csv_. This file contains counts for the number of three different species of lizards found at 5 different locations in a region.
If we inspect the data we see that it is in a long format.

```{r}
lizard <- read.csv("lizard-workshop1.csv")
print(lizard)
```

Each row represent a species and each column a location. This is not a helpful format to use in R, as many functions will be very difficult to run. For instance, if we wanted to plot the number of lizard in each species by location we don't have a straightforward way of doing it ^[There are ways of plotting wide data, of course, but for instance functions such as _lm_ want a long data format.]

Luckily R comes to the rescue! For reshaping data we need to install an extra R package, called _reshape2_ ^[There are also other packages such as _tidyr_ that can perform the same task, feel free to explore those as well if you wish.]. To install the package we can just run:

```{r eval=F}
install.packages("reshape2")
```

You need to do this only once, if you already have the package installed the line above is unnecessary. Once the package is installed we need to tell R to use it; we do so by using the _library_ command.

```{r}
library(reshape2)
```

We can now use all the functions provided by _reshape2_, specifically we want the _melt_ function, that converts data from wide to long.

```{r}
lizard.long <- melt(lizard, id.vars = "Lizard", variable.name = "Location", value.name = "Counts")
```

This tells R to use column Lizard as our _ID_, that is the identifier of the subject that is being measured. The _variable.name_ parameters specifies the name of the variable(s)
 that we are changing, in this case the location (this is just the name of the new column in the long dataset). Finally, value.name is the column name for what we are currently measuring.
 
Let's see what we got!

```{r}
head(lizard.long)
```
\newpage
We can now easily do:

```{r}
boxplot(Counts ~ Location, lizard.long, las=2, ylab = "Number of lizards")
boxplot(Counts ~ Lizard, lizard.long, las=2, ylab = "Number of lizards")
```

Sometimes conversions from wide to long data are a bit more involved, but _melt_ is a very powerful function and most of the time you will be able to reshape your data.

\newpage
Now try to open the file _BP-workshop1.csv_. This contains data from an experiment where participants of different sex and age took two different drugs. 
The data has been anonymised, so that each participant is represented by an ID (2 letters and 3 numbers). Each participant blood pressure was measured before and after taking each of the drugs, and the relative change in blood pressure was calculated. The dataset contains the calculated fold changes for each participants.

- In which format is the data (e.g. wide, long, ...) ? 
- If necessary, convert the data into long format 
- How many participants were in the study? How many men and how many women?
- How many participants were over 50 years old? How many of those were women?
- Is the response to the drugs different between men and women? What about between younger and older participants?
- Plot the response to the two drugs using an histogram, a boxplot, or a barplot. Which representation is more useful? Why?
- Look at the data for patient PY080. Is there anything striking about this patient? What do you think you should do if you were to continue and analyse these data?

Finally, read the file _dataset3.csv_ and answer these questions:

1. xxx
2. xxx
3. xxx
4. xxx
5. xxx